---
title: "Introduction to LODopt"
author: "Reuben Thomas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to LODopt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(LODopt)
library(dplyr)
library(magrittr)
library(SummarizedExperiment)
```

## Introduction

The `LODopt` package provides tools for analyzing associations between 
cell type abundances and experimental conditions in single-cell RNA-seq data.

## Quick Start

### Example Data
We simulate data for 30 samples with 25 clusters, where 4 clusters (c0, c2, c7 and c14) have changed their abundance. 
```{r simulate-counts}
# Create example count matrix
  # Create test data
  set.seed(123)
   # No of clusters in single cell dataset
   K=25
   # No of cells in tissue
   depth = 1e9
   # Total number of samples
   nsamp = 30
   alpha = 10^runif(K, min=log10(0.5), max = log10(10))
   p <- dirmult::rdirichlet(alpha = alpha) |> sort()
   p <- p[p > 0.001]
   p <- p/sum(p)
   size <- rep(10, length(p))
   change_mean = rep(1, length(p))
   ##changed cluster indices
   change_mean[c(1,3,8,15)] = c(0.2, 2, 0.2, 2)
   # Simulate counts
   counts_res <- simulate_cellCounts_fromTissue(props=p,nsamp=nsamp,depth=depth, size = size, change_mean = change_mean)
```

### Set up the required SummarizedExperiment object
```{r}
  counts <- counts_res$counts

  pheno_data <- data.frame(sampleID = paste0("S", 1:30),
                           groupid = c(rep("group0", 15), rep("group1", 15)))
  pheno_data %<>% tibble::column_to_rownames("sampleID")



  require(SummarizedExperiment)
  model_formula <- "groupid"
  cellcomp_se <- SummarizedExperiment(assays = list(counts=counts),
                                      colData = pheno_data,
                                      metadata = list(modelFormula = model_formula,
                                                      coef_of_interest_index = 2,
                                                      reference_levels_of_variables = list(c("groupid", "group0")),
                                                      random_seed = 123456,
                                                      unchanged_cluster_indices = NULL))

```
### Running Association Analysis
```{r}
cellcomp_res <- logodds_optimized_normFactors(cellcomp_se)
```

### Summarizing Results
The estimate column has the log odds ratio estimates for the cluster-specific association with groupid while the estimates_significance has the corresponding raw p-values for the corresponding null hypothesis that the log odds ratio is equal to 0.
```{r}
print(cellcomp_res$res)
```



## Conclusion

The `LODopt` package provides a simple interface for testing associations
between cell type abundances and experimental conditions, with support for 
multiple statistical methods and comprehensive result visualization.
